---
title: "Using mdbplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using mdbplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This repository contains `mdbplyr`, an R package that provides a disciplined,
lazy `dplyr`-style interface for MongoDB aggregation pipelines.

`mdbplyr` sits between raw `mongolite` usage and broader compatibility layers.
Compared with writing aggregation JSON by hand, it lets you express supported
queries with tidy verbs and inspect the generated pipeline before execution.
Compared with approaches that try to hide MongoDB behind table-like semantics,
it stays explicit about scope, keeps translation conservative, and fails clearly
when a verb or expression is outside the supported subset.

The practical advantages are:

- lazy query construction until `collect()`,
- cursor-based streaming with `cursor()`,
- direct translation to MongoDB aggregation stages,
- `show_query()` for inspectable pipeline output,
- explicit unsupported-feature errors instead of silent fallback,
- a smaller, testable semantic surface aligned with MongoDB rather than SQL.

# Usage examples

## Loading data

Suppose there is a running MongoDB instance on `localhost` with default port
and no authentication. The code below loads `dplyr::starwars` into a collection
named `starwars`.

```{r load-starwars, eval = FALSE}
library(dplyr)
library(mongolite)
library(mdbplyr)

starwars_collection <- mongolite::mongo(
  collection = "starwars",
  db = "mdbplyr"
)

starwars_collection$drop()
starwars_collection$insert(dplyr::starwars)

starwars_tbl <- tbl_mongo(
  starwars_collection,
  schema = names(dplyr::starwars)
)
```

Once the collection is loaded, `starwars_tbl` is the lazy table used in the
examples below.

```{r create-lazy-table, eval = FALSE}
library(dplyr)
library(mdbplyr)

starwars_collection <- mongolite::mongo(
  collection = "starwars",
  db = "mdbplyr"
)

starwars_tbl <- tbl_mongo(
  starwars_collection,
  schema = names(dplyr::starwars)
)
```

## Basic inspection

Inspect the known schema and the generated pipeline without executing the
query.

```{r inspect, eval = FALSE}
schema_fields(starwars_tbl)

starwars_tbl |>
  filter(species == "Human", height > 180) |>
  select(name, height, mass) |>
  show_query()
```

## Streaming with `cursor()`

When you want a MongoDB iterator instead of an eagerly materialized tibble,
open the lazy query with `cursor()`.

```{r inspect-cursor, eval = FALSE}
iter <- starwars_tbl |>
  filter(species == "Human", height > 180) |>
  select(name, height, mass) |>
  cursor()

iter$page(10)
```

# Supported verbs

Each subsection below shows one of the supported `dplyr`-like verbs on the
`starwars` collection.

## `filter()`

```{r verb-filter, eval = FALSE}
starwars_tbl |>
  filter(species == "Droid", height > 100) |>
  collect()
```

## `select()`

```{r verb-select, eval = FALSE}
starwars_tbl |>
  select(name, species, homeworld) |>
  collect()
```

## `rename()`

```{r verb-rename, eval = FALSE}
starwars_tbl |>
  rename(character_name = name, planet = homeworld) |>
  collect()
```

## `mutate()`

```{r verb-mutate, eval = FALSE}
starwars_tbl |>
  mutate(height_m = height / 100, bmi_like = mass / (height_m * height_m)) |>
  select(name, height, mass, height_m, bmi_like) |>
  collect()
```

## `transmute()`

```{r verb-transmute, eval = FALSE}
starwars_tbl |>
  transmute(name = name, height_m = height / 100) |>
  collect()
```

## `arrange()`

```{r verb-arrange, eval = FALSE}
starwars_tbl |>
  arrange(desc(height), name) |>
  select(name, height) |>
  slice_head(n = 10) |>
  collect()
```

## `group_by()`

```{r verb-group-by, eval = FALSE}
starwars_tbl |>
  group_by(species)
```

## `summarise()`

```{r verb-summarise, eval = FALSE}
starwars_tbl |>
  group_by(species) |>
  summarise(
    n = n(),
    avg_height = mean(height),
    max_mass = max(mass)
  ) |>
  arrange(desc(n)) |>
  collect()
```

## `slice_head()`

```{r verb-slice-head, eval = FALSE}
starwars_tbl |>
  select(name, species) |>
  slice_head(n = 5) |>
  collect()
```

## `head()`

```{r verb-head, eval = FALSE}
head(starwars_tbl, 5) |>
  collect()
```

# Join verbs

`inner_join`, `left_join`, `semi_join`, and `anti_join` are supported for
`tbl_mongo` objects.  They compile lazily to MongoDB `$lookup` aggregation
stages using the **correlated-pipeline form** (MongoDB 3.6+), so:

- multiple join keys are handled natively,
- any filters or transformations already applied to the right-hand table
  are embedded inside the `$lookup` pipeline,
- the full join stays lazy until `collect()`.

Both the left and right table must be `tbl_mongo` objects backed by
collections in the **same** MongoDB database.

`right_join()` and `full_join()` are not supported because MongoDB has no
native full outer join stage starting from the left collection's pipeline.
`right_join` can always be rewritten as `left_join` with the tables swapped.

## Setup for join examples

The examples below use two collections: `orders` and `customers`.

```{r join-setup, eval = FALSE}
orders_collection <- mongolite::mongo("orders", db = "mdbplyr")
customers_collection <- mongolite::mongo("customers", db = "mdbplyr")

# Insert sample data
orders_collection$drop()
orders_collection$insert(
  data.frame(order_id = 1:4, customer_id = c(1L, 2L, 1L, 5L), amount = c(10, 20, 30, 40))
)

customers_collection$drop()
customers_collection$insert(
  data.frame(customer_id = 1:3, name = c("Alice", "Bob", "Charlie"), country = c("Italy", "France", "Italy"))
)

orders_tbl    <- tbl_mongo(orders_collection,    schema = c("order_id", "customer_id", "amount"))
customers_tbl <- tbl_mongo(customers_collection, schema = c("customer_id", "name", "country"))
```

## `inner_join()`

Returns only rows that have a matching record in both tables.

```{r verb-inner-join, eval = FALSE}
inner_join(orders_tbl, customers_tbl, by = "customer_id") |>
  collect()
```

## `left_join()`

Returns all rows from the left table; columns from the right table are `NA`
where there is no match.

```{r verb-left-join, eval = FALSE}
left_join(orders_tbl, customers_tbl, by = "customer_id") |>
  collect()
```

## `semi_join()`

Returns rows from the left table that have at least one match in the right
table. No columns from the right table are added.

```{r verb-semi-join, eval = FALSE}
semi_join(orders_tbl, customers_tbl, by = "customer_id") |>
  collect()
```

## `anti_join()`

Returns rows from the left table that have **no** match in the right table.

```{r verb-anti-join, eval = FALSE}
anti_join(orders_tbl, customers_tbl, by = "customer_id") |>
  collect()
```

## Inspecting the generated pipeline

Use `show_query()` to inspect the MongoDB aggregation pipeline that will be
sent to the server.

```{r join-show-query, eval = FALSE}
inner_join(orders_tbl, customers_tbl, by = "customer_id") |>
  show_query()
```

## Pre-filtering the right-hand table

Because join compilation is lazy, any filters already applied to `y` are
embedded inside the `$lookup` sub-pipeline and executed server-side.

```{r join-prefilter, eval = FALSE}
italian_customers <- filter(customers_tbl, country == "Italy")

inner_join(orders_tbl, italian_customers, by = "customer_id") |>
  collect()
```

## Multi-key joins

Pass a named vector to `by` when the key columns have different names in each
table, or an unnamed vector when the names match.

```{r join-multikey, eval = FALSE}
# Same key name in both tables
inner_join(orders_tbl, customers_tbl, by = "customer_id") |> collect()

# Different key names: left column "oid" matches right column "order_id"
inner_join(orders_tbl, other_tbl, by = c("order_id" = "inv_order_id")) |> collect()
```

# Notes

The examples above stay within the currently supported subset:

- explicit field references,
- scalar expressions for `mutate()` and `transmute()`,
- conservative grouped summaries,
- `inner_join`, `left_join`, `semi_join`, `anti_join` via MongoDB `$lookup`,
- no window functions or automatic client-side fallback.

Join support requires MongoDB 3.6 or later (correlated-pipeline `$lookup`).
Both tables must be backed by collections in the same database.

If a query falls outside that subset, `mdbplyr` is designed to fail
explicitly rather than guess or silently change execution semantics.
