---
title: "Using mdbplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using mdbplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This repository contains `mdbplyr`, an R package that provides a disciplined,
lazy `dplyr`-style interface for MongoDB aggregation pipelines.

`mdbplyr` sits between raw `mongolite` usage and broader compatibility layers.
Compared with writing aggregation JSON by hand, it lets you express supported
queries with tidy verbs and inspect the generated pipeline before execution.
Compared with approaches that try to hide MongoDB behind table-like semantics,
it stays explicit about scope, keeps translation conservative, and fails clearly
when a verb or expression is outside the supported subset.

The practical advantages are:

- lazy query construction until `collect()`,
- cursor-based streaming with `cursor()`,
- direct translation to MongoDB aggregation stages,
- `show_query()` for inspectable pipeline output,
- explicit unsupported-feature errors instead of silent fallback,
- a smaller, testable semantic surface aligned with MongoDB rather than SQL.

# Usage examples

## Loading data

Suppose there is a running MongoDB instance on `localhost` with default port
and no authentication. The code below loads `dplyr::starwars` into a collection
named `starwars`.

```{r load-starwars, eval = FALSE}
library(dplyr)
library(mongolite)
library(mdbplyr)

starwars_collection <- mongolite::mongo(
  collection = "starwars",
  db = "mdbplyr"
)

starwars_collection$drop()
starwars_collection$insert(dplyr::starwars)

starwars_tbl <- tbl_mongo(
  starwars_collection,
  schema = names(dplyr::starwars)
)
```

Once the collection is loaded, `starwars_tbl` is the lazy table used in the
examples below.

```{r create-lazy-table, eval = FALSE}
library(dplyr)
library(mdbplyr)

starwars_collection <- mongolite::mongo(
  collection = "starwars",
  db = "mdbplyr"
)

starwars_tbl <- tbl_mongo(
  starwars_collection,
  schema = names(dplyr::starwars)
)
```

## Basic inspection

Inspect the known schema and the generated pipeline without executing the
query.

```{r inspect, eval = FALSE}
schema_fields(starwars_tbl)

starwars_tbl |>
  filter(species == "Human", height > 180) |>
  select(name, height, mass) |>
  show_query()
```

## Streaming with `cursor()`

When you want a MongoDB iterator instead of an eagerly materialized tibble,
open the lazy query with `cursor()`.

```{r inspect-cursor, eval = FALSE}
iter <- starwars_tbl |>
  filter(species == "Human", height > 180) |>
  select(name, height, mass) |>
  cursor()

iter$page(10)
```

# Supported verbs

Each subsection below shows one of the supported `dplyr`-like verbs on the
`starwars` collection.

## `filter()`

```{r verb-filter, eval = FALSE}
starwars_tbl |>
  filter(species == "Droid", height > 100) |>
  collect()
```

## `select()`

```{r verb-select, eval = FALSE}
starwars_tbl |>
  select(name, species, homeworld) |>
  collect()
```

## `rename()`

```{r verb-rename, eval = FALSE}
starwars_tbl |>
  rename(character_name = name, planet = homeworld) |>
  collect()
```

## `mutate()`

```{r verb-mutate, eval = FALSE}
starwars_tbl |>
  mutate(height_m = height / 100, bmi_like = mass / (height_m * height_m)) |>
  select(name, height, mass, height_m, bmi_like) |>
  collect()
```

## `transmute()`

```{r verb-transmute, eval = FALSE}
starwars_tbl |>
  transmute(name = name, height_m = height / 100) |>
  collect()
```

## `arrange()`

```{r verb-arrange, eval = FALSE}
starwars_tbl |>
  arrange(desc(height), name) |>
  select(name, height) |>
  slice_head(n = 10) |>
  collect()
```

## `group_by()`

```{r verb-group-by, eval = FALSE}
starwars_tbl |>
  group_by(species)
```

## `summarise()`

```{r verb-summarise, eval = FALSE}
starwars_tbl |>
  group_by(species) |>
  summarise(
    n = n(),
    avg_height = mean(height),
    max_mass = max(mass)
  ) |>
  arrange(desc(n)) |>
  collect()
```

## `slice_head()`

```{r verb-slice-head, eval = FALSE}
starwars_tbl |>
  select(name, species) |>
  slice_head(n = 5) |>
  collect()
```

## `head()`

```{r verb-head, eval = FALSE}
head(starwars_tbl, 5) |>
  collect()
```

# Notes

The examples above stay within the currently supported subset:

- explicit field references,
- scalar expressions for `mutate()` and `transmute()`,
- conservative grouped summaries,
- no joins, window functions, or automatic client-side fallback.

If a query falls outside that subset, `mdbplyr` is designed to fail
explicitly rather than guess or silently change execution semantics.
